Increased productivity using Atomify - a real-time LAMMPS visualizer

### 2018-06-05 ms.tex
Fixed simulation and intro language

Thankfully, there already exists several software packages for atomistic simulations, many of which have been developed for more than two decades\footnote{\url{https://en.wikipedia.org/wiki/Comparison_of_software_for_molecular_mechanics_modeling} compares different softwares for molecular modeling.}.
Some of these packages, such as LAMMPS\citep{Plimpton1995Fast}, Gromacs\citep{berendsen1995gromacs}, NAMD\citep{Phillips2005Scalable} and CHARMM\citep{brooks2009charmm}, are open source and being actively developed with hundreds of contributers.
%After the rise of powerful graphical processing units (GPU's), most of these codes eventually added support of using accelerators\citep{brown2011implementing}.

@@

The work in this paper is focused on LAMMPS.
Although the software is well-documented\footnote{The LAMMPS documentation is known for being very detailed and precise.} and enables the user to perform advanced simulations, it can be demanding to learn how to use it well.
The core principles of molecular dynamics is so simple that high school students can easily understand the concepts, but even undergraduate students may have problems with the terminology of statistical mechanics which is used in several of the commands.
For instance, to simulate an \textit{NVE} simulation (constant number of particles $N$, volume $V$ and energy $E$), we use the \code{fix nve} command.

@@

If we want to measure physical quantities such as energy, temperature and stress, they can be accessed through a series of commands called \textit{computes}.
For instance, the commands \code{compute ke} and \code{compute pe} measures the kinetic and the potential energy.
Using \textit{fixes}, we can create time averages and save these quantities to one or more files for plotting and further analysis.
We also often put atoms in different \textit{groups}, either by their position being in a \textit{region}, or their atom type or more advanced criteria.
Performing different operations on different groups, or just measuring physical quantities on a specific group is then easily done since almost all commands have \code{group} as an input parameter.

@@

The lack of an intuitive GUI and visualization\footnote{GUI is listed as a non-feature on the LAMMPS web page: \url{http://lammps.sandia.gov/non_features.html}.}
makes the workflow complicated, see Fig. \ref{fig:flowchart}.
A simulation with full analysis usually involves writing the script in a text editor, running the simulation using the terminal before analysis is done with a visualization program and additional scripts written in MATLAB or Python.
When creating a new simulation, the initial state (system box, positions of all particles, groups and regions) can be tricky to set up right, especially since groups and regions are not automatically written to the output file through the \code{dump} command.
The time from an idea to a visual feedback can then be destructive for the creativity where immediate feedback would be ideal.

@@

The visualization is performed using the powerful Qt3D library where advanced rendering techniques such as billboard raytracing is used to achieve maximum performance.
This allows visualization of millions of atoms with good framerate on modern desktop computers.
Atomify is designed with simplicity as its main goal.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{gui.png}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar (1) is used to change between the different modes, such as 
    editing, viewing and analyzing the running simulation, 
    in addition to playback controls.
    The file browser (2) contain the list of open LAMMPS scripts or data files.
    The text editor (3) provides code highlighting.
    The viewport (4) shows the current simulation.
    The simulation properties (5) shows the simulations details, gives access to 
    LAMMPS objects such as groups, regions and computes, and the rendering
    properties (6) lets you change the viewport settings, such as lighting and
    draw mode.
    TODO: Add numbers to figure.
    }
	\label{fig:gui}
\end{figure*}

@@

\section{\label{sec:casestudy}Case study: flow through nanoporous media}
\todo{Short motivation about flow in nanoporous media with references}
Fluid flow in porous media is often characterized with permeability, the inverse flow resistance of a material.
The permeability can be measured using Darcy's law when applying a pressure gradient.
However, the derivation of Darcy's law assume no-slip boundary conditions, i.e. the fluid velocity is zero at the boundary of a channel. (ref)
This is a good approximation for liquids in large pores, but if the channel diameter $L$ is similar or smaller than the mean free path of the fluid $\lambda$, this is no longer the case.
The Knudsen number, $\text{Kn} = \lambda / L$ quantifies this and can be used with the Klinkenberg correction\citep{klinkenberg1941permeability} to correctly describe the effective permeability as a function of pressure in the high Knudsen number limit.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{lj_flow/configuration.png}
	\caption{
		Initial configuration in our case study simulation.
		A periodic box consisting of $(40\times20\times20)$ FCC unit cells with reduced density $\rho^* = 0.8442$ where we will have gas flow in a cylinder of 5 unit cell radius.
		The different colors indicate the regions of which different integration rules apply.
		\textcolor{atomify-blue}{Blue} atoms are fixed and do not move at all,
		whereas the \textcolor{atomify-yellow}{yellow} atoms are thermalized with a Langevin\citep{schneider1978molecular} dissipative thermostat to mimick the heat exchange with the frozen bulk atoms.
		The \textcolor{atomify-red}{red} and the \textcolor{atomify-green}{green} atoms are integrated normally, except the \textcolor{atomify-green}{green} gas atoms have an applied constant acceleration representing the pressure gradient.
		Setting up and verifying this system is simple using Atomify due to the real-time coloring of groups and regions.
    }
	\label{fig:cylinder_simulation}
\end{figure}

In this case study, we will use Atomify to measure the velocity profile of a Lennard Jones gas inside a nanotube with varying density to observe the slip velocity effect at low densities.
We will create the nanotube inside a Lennard Jones solid of size  $(40\times20\times20)$ unit cells, reduced density $\rho^* = 0.8442$, where the fluid will flow due to an induced pressure gradient, see Fig. \ref{fig:cylinder_simulation}.
Here we apply a pressure gradient $\nabla P$ in the $x$-direction on the gas atoms using a constant acceleration $a$ calculated as
\begin{align}
	a = \frac{\nabla P}{\rho_m},
\end{align}
where $\rho_m$ is the mass density of the gas atoms.
Applying such an acceleration means adding energy to the system which in turn will increase the temperature and eventually melt the system.
We want to couple the solid to an external heat bath so the added energy can be transferred out through a thermostat.
Even though we only apply a pressure gradient on the gas atoms, this induced momentum will also be transferred to the solid atoms which eventually will start moving in the positive $x$-direction which is not desired.
We therefore freeze the outmost atoms.
To achieve these properties of the system, the atoms are divided into multiple regions, which are integrated differently:

\begin{itemize}  
	\item $r \leq 5$: applied pressure gradient
	\item $5 < r \leq 9$: regular time integration
	\item $9 < r \leq 12$: Dissipative Langevin thermostat
	\item $12 < r$: frozen.
\end{itemize}

\subsection{Creating the initial geometry}
We create a new script in Atomify using \keys{\cmd+N}, and save it (\keys{\cmd+S}) as \code{run.in}.
Starting from the input script in listing \ref{lst:simple.in}, we modify it to create a larger simulation box with $(40\times20\times20)$ unit cells.
Remove the line \code{fix nve all nve} so atoms will stand still (we will add integrator later anyway).
We will modify this script to achieve what we described above, so we need to make sure that these commands happen \textit{before} the \code{run 1000} command in the script.
If not, LAMMPS will not execute them before we start visualizing.
To define the inner cylinder, we first specify the radius and the center before using the region command
\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	variable R equal $(ly*0.25)
	variable c equal $(ly*0.5)
	region cylinder1 cylinder x $c $c $R EDGE EDGE
\end{lstlisting}
The region command will create a cylinder in the $x$-direction, with $(y,z)$-coordinates at the system center \code{\$c} with radius \code{\$R} using the full length of the cylinder specified when using \code{EDGE}.
We now press \keys{\cmd+R} to run the simulation to see how it looks.
As shown in Fig. \ref{fig:rightbar}, we can hover the specified region to highlight the atoms in that region.
When we hover the region, we immediately see that the cylinder is in fact not placed at the center of the system, as shown in Fig. \ref{fig:initial_configuration}a.
The reason is that LAMMPS operate with two different units, either the length unit or the lattice length unit. This is explained in the documentation, but is a common mistake.
To discover mistakes like this is quickly done with Atomify.
Since we have used \code{\$(0.5*lx)}, we need to specify \code{units box} in the region command, rerun the simulation with \keys{\cmd+R} which gives the desired result in figure  \ref{fig:initial_configuration}b.

>>>>

Thankfully, there already exists several software packages for atomistic simulations, many of which have been developed for more than two decades\footnote{\url{https://en.wikipedia.org/wiki/Comparison_of_software_for_molecular_mechanics_modeling} compares different softwares for molecular modeling.}.
Some of these packages, such as LAMMPS\citep{Plimpton1995Fast}, Gromacs\citep{berendsen1995gromacs}, NAMD\citep{Phillips2005Scalable} and CHARMM\citep{brooks2009charmm}, are open source and being actively developed with hundreds of contributers\footnote{See their repositories on i.e. GitHub for an overview of contributers.}.
%After the rise of powerful graphical processing units (GPU's), most of these codes eventually added support of using accelerators\citep{brown2011implementing}.

@@

The work in this paper is focused on LAMMPS.
Although the software is well-documented\footnote{The LAMMPS documentation is known for being very detailed and precise.}, and enables the user to perform advanced simulations, it can be a demanding task to learn how to use it well.
The core principles of molecular dynamics is so simple that high school students can easily understand the concepts, but even undergraduate students may have problems with the terminology of statistical mechanics which is used in several of the commands.
For instance, to simulate an \textit{NVE} simulation (constant number of particles $N$, volume $V$ and energy $E$), we use the \code{fix nve} command which defines the time integration.

@@

If we want to measure physical quantities such as energy, temperature and stress, they can be accessed through commands called \textit{computes}.
For instance, the commands \code{compute ke} and \code{compute pe} measures the kinetic and the potential energy.
Using \textit{fixes}, we can create time averages and save them to one or more files for plotting and further analysis.
We also often put atoms in different \textit{groups}, either by their position being in a \textit{region}, or their atom type or more advanced criteria.
Performing different operations on different groups, or just measuring physical quantities on a specific group is then easily done since almost all commands have \code{group} as an input parameter.
With this flexibility, it may be difficult to do everything right immediately.
Complicated regions and group defines are often necessary to perform the intended simulation, and a lot of time can be spent in the prototyping process, especially since information about groups and regions are not automatically written to a trajectory file.
A simulation typically involves writing the LAMMPS script in a text editor before the simulation is performed in the terminal for a while before the trajectories and time series are studied to verify the correctness which often turn out to be bad.
Plotting of time series also requires the user to write a script in i.e. Python or MATLAB which does the plotting, but these steps are often performed manually.

@@

The lack of an intuitive GUI and visualization\footnote{GUI is listed as a non-feature on the LAMMPS web page: \url{http://lammps.sandia.gov/non_features.html}.}
makes this workflow complicated, see Fig. \ref{fig:flowchart}.
The time from an idea to a visual feedback can then be destructive for the creativity where immediate feedback would be ideal.

@@

The visualization is performed using the powerful Qt3D library where we have implemented advanced rendering techniques such as billboard raytracing to achieve maximum performance.
This allows visualization of millions of atoms with good framerate on modern desktop computers.
Atomify is designed with simplicity as its main goal.

@@

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{gui.png}
	\caption{%
    Overview of the Graphical User Interface (GUI) in Atomify.
    The toolbar (1) is used to change between the different modes, such as 
    editing, viewing and analyzing the running simulation, 
    in addition to playback controls.
    The file browser (2) contain the list of open LAMMPS scripts or data files.
    The text editor (3) provides code highlighting.
    The viewport (4) shows the current simulation.
    The simulation properties (5) shows the simulations details, gives access to 
    LAMMPS objects such as groups, regions and computes, and the rendering
    properties (6) lets you change the viewport settings, such as lighting and
    draw mode.
    TODO: Add numbers to figure.
    }
	\label{fig:gui}
\end{figure*}

@@

\section{\label{sec:casestudy}Case study: flow through nanoporous media}
The features of Atomify is best presented through a case study where many of the features are useful to be productive in the script development.
We have chosen to create a system with fluid flow inside a tight nanochannel.
This involves creating a somewhat complicated initial geometry with multiple regions where we apply different tricks such as dissipative thermostats, constant force mimicking applied pressure gradient and frozen atoms to avoid net momentum of the solid.

Fluid flow in porous media is often characterized with permeability, the inverse flow resistance of a material.
The permeability can be measured using Darcy's law when applying a pressure gradient.
However, the derivation of Darcy's law assume no-slip boundary conditions, i.e. the fluid velocity is zero at the boundary of a channel. (ref)
This is a good approximation for liquids in larger pores, but if the channel size $R$ is of the same order, or larger, as the mean free path $\lambda$ of the fluid, this is no longer true. (ref)
The Knudsen number, $\text{Kn} = \lambda / R$ quantifies this and can be used to apply Klinkenberg correction\citep{klinkenberg1941permeability} to correctly describe the effective permeability as a function of pressure in the high Knudsen number limit.

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{lj_flow/configuration.png}
	\caption{
		Initial configuration in our case study simulation.
		A periodic box consisting of $(40\times25\times25)$ FCC unit cells with reduced density $\rho^* = 0.8442$ where we will have gas flow in a cylinder with radius 10 (reduced length).
		The different colors indicate the regions of which different integration rules apply.
		\textcolor{atomify-blue}{Blue} atoms are fixed and do not move at all,
		whereas the \textcolor{atomify-yellow}{yellow} atoms are thermalized with a Langevin\citep{schneider1978molecular} dissipative thermostat to mimick the heat exchange with the frozen bulk atoms.
		The \textcolor{atomify-red}{red} and the \textcolor{atomify-green}{green} atoms are integrated normally, except the \textcolor{atomify-green}{green} gas atoms have an applied constant acceleration representing the pressure gradient.
		Setting up and verifying this system is simple using Atomify due to the real-time coloring of groups and regions.
    }
	\label{fig:cylinder_simulation}
\end{figure}

In this case study, we will use Atomify to measure the velocity profile of a Lennard Jones gas inside a nanotube with varying density to observe the slip velocity effect at low densities.
We will create the nanotube inside a Lennard Jones solid of size  $(40\times25\times25)$ unit cells with reduced density $\rho^* = 0.8442$. The fluid in the nanochannel, or cylinder, will flow due to an induced pressure gradient, see Fig. \ref{fig:cylinder_simulation}.
Here we apply a pressure gradient $\nabla P$ in the $x$-direction on the gas atoms using a constant acceleration $a$ calculated as
\begin{align}
	a = \frac{\nabla P}{\rho_m},
\end{align}
where $\rho_m$ is the mass density of the gas atoms.
Applying such an acceleration means adding energy to the system which in turn will increase the temperature and eventually melt the system.
To avoid this, we will couple the solid to an external heat bath so the added energy can be transferred away from the system through a thermostat.
Even though we only apply a pressure gradient on the gas atoms, this induced momentum will also be transferred to the solid atoms which eventually will start moving in the positive $x$-direction which is not desired.
We therefore freeze the outmost atoms (blue in Fig \ref{fig:cylinder_simulation}).
To achieve these properties of the system, the atoms are divided into multiple regions, which are integrated differently:

\begin{itemize}  
	\item $r \leq 10$: applied pressure gradient
	\item $5 < r \leq 4$: regular time integration
	\item $9 < r \leq 12$: Dissipative Langevin thermostat
	\item $12 < r$: frozen.
\end{itemize}
The different radii are in reduced Lennard Jones units.

\subsection{Creating the initial geometry}
We create a new script in Atomify using \keys{\cmd+N}, and save it (\keys{\cmd+S}) as \code{run.in}.
Starting from the input script in listing \ref{lst:simple.in}, we modify it to create a larger simulation box with $(40\times25\times25)$ unit cells.
Remove the line \code{fix nve all nve} so atoms will stand still (we will add integrator later anyway).
We will modify this script to achieve what we described above, so we need to make sure that these commands happen \textit{before} the \code{run 1000} command in the script.
If not, LAMMPS will not execute them before we start visualizing.
To define the inner cylinder, we first specify the radius and the center before using the region command
\begin{lstlisting}[basicstyle=\tiny, frame = none, numbers=none, framexleftmargin=0pt, xleftmargin=-0.75cm, xrightmargin=0.0cm]
	variable R equal $(ly*0.25)
	variable c equal $(ly*0.5)
	region cylinder1 cylinder x $c $c $R EDGE EDGE
\end{lstlisting}
The region command will create a cylinder in the $x$-direction, with $(y,z)$-coordinates at the system center \code{\$c} with radius \code{\$R} using the full length of the cylinder specified when using \code{EDGE}.
We now press \keys{\cmd+R} to run the simulation to see how it looks.
As shown in Fig. \ref{fig:rightbar}, we can hover the specified region to highlight the atoms in that region.
When we hover the region, we immediately see that the cylinder is in fact not placed at the center of the system, as shown in Fig. \ref{fig:initial_configuration}a.
The reason is that LAMMPS operate with two different units, either the length unit or the lattice length unit. This is explained in the documentation, but is a common mistake.
To discover mistakes like this is quickly done with Atomify.
Since we have used \code{\$(0.5*lx)}, we need to specify \code{units box} in the region command, rerun the simulation with \keys{\cmd+R} which gives the desired result in figure  \ref{fig:initial_configuration}b.
